Modelling state-switching functional data with hidden Markov models

### 2020-11-06 Report/Sections/appendix.tex
Updates from Overleaf

It is straightforward to simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$. 
To simulate $X^*_t$ for each dive $t$, note that we simulate accelerometer readings taken at 50 hertz and have two-second window lengths. Therefore, each window contains 100 accelerometer readings and $X^*_t = \left\{X^*_{t,1}, X^*_{t,101}, \ldots, X^*_{t,100T^*_t-99}\right\}$. In addition, the length of each fine-scale Markov chain is $T^*_t = \lfloor Y_t/2 \rfloor$. Then, given the dive types $X_t$, dive durations $Y_t$, and the fine-scale probability transition matrices $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$, it is straightforward to simulate $X^*_t$ for each dive $t = 1,2,\ldots,100$. Recall that $\Gamma^{*(i)}_{jk} = Pr(X^*_{t,t^*+100} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1, 101, \ldots$. It is much more difficult to simulate the raw acceleration data $Y^*$, and we focus on this task for the remainder of this section.

Assume that $X$, $Y$, and $X^*$ are simulated using the procedure described above and are therefore known. Additionally, let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for convenience. We first simulate $\hat Y^{*(k)}_{t,t^*}$ for all dives $t = 1,2,\ldots,100$, all subdive windows $t^* = 1,101,\ldots$, and all Fourier coefficients $k = 0,1,\ldots,99$. We then take the inverse DFT of $\hat{Y}^*_{t,t^*}$ for all dives $t = 1,2,\ldots,100$ and subdive windows $t^* = 1,101,\ldots$ to find $Y^*$. The details of this procedure are described below:

>>>>

It is straightforward to simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$. We then simulate simulate $X^*_t$ for a fixed dive $t$. Since accelerometer readings are taken at 50 hertz and this model has two-second window lengths, each window contains 100 accelerometer readings. In addition, the length of each fine-scale Markov chain is $\tilde{T}^*_t = \lfloor Y_t/2 \rfloor$. Then, given the dive type $X_t = i$, dive duration $Y_t = y_t$, and the fine-scale probability transition matrix $\Gamma^{*(i)}$, it is straightforward to simulate the subdive states $\tilde{X}^*_t$. It is much more difficult to simulate the raw acceleration data $Y^*$, and we focus on this task for the remainder of this section.

Let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for convenience. We first simulate $\hat Y^{*(k)}_{t,t^*}$ for all subdive windows $t^* = 1,101,\ldots$, and all Fourier coefficients $k = 0,1,\ldots,99$. We then take the inverse DFT of $\hat{Y}^*_{t,t^*}$ for all subdive windows $t^* = 1,101,\ldots$ to find $Y^*$. The details of this procedure are described below:

### 2020-11-06 Report/Sections/results.tex
Updates from Overleaf

We then select a model corresponding to the fine-scale observations of acceleration. Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$. Although $N^*$ is selected heuristically, we test the validity of this model using techniques in Section \ref{subsec:model_validation}.
%
In contrast to the coarse-scale observations, the fine-scale acceleration data exhibit significant sinusoidal behaviour. Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with window size $h=100$ (two seconds) and maximum frequency $\tilde{\omega}=10$ (5 Hz). 
%We select a window size of two seconds since it appears be the smallest window size that captures at least one full period of each ``wiggle" in the acceleration data. 
We then have that $\z_{t,t^*} = \left\{\zone_{t,t^*},\ztwo_{t,t^*}\right\}$, where $\zone_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,t^*}$ is a scalar describing the ``wiggliness" of a particular window. We also down-sample each observation $\z_t$ such that $\z_t = \left\{\z_{t,1},\z_{t,101},\z_{t,201},\ldots\right\}$ and $x^*_t = \left\{x^*_{t,1},x^*_{t,101},x^*_{t,201},\ldots\right\}$ to avoid overlapping windows. Even after transforming and down-sampling the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material). Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.

We then select the specific emission distribution of $\Z_{t,t^*}$. Given the dive types and subdive states, we assume that $\Ztwo_{t,t^*}$ and all components of $\Zone_{t,t^*}$ are independent of one another. To reduce model complexity, we also assume that the three fine-scale emission parameters are shared across the two dive types $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$. This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.

All components of $\Zone_{t,t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) for all $t$ and $t^*$. Each component is assumed to have its own mean and variance parameters, but all components have the same auto-correlation parameter. 
%Given that $X^*_{t,t^*} = i^*$, the corresponding unknown parameters are then $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}_{>0}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
Given that $X^*_{t,t^*} = i^*$, $\Ztwo_{t,t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$. We do not include $\Ztwo_{t,t^*-100}$ in the distribution of $\Ztwo_{t,t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state. 

@@

%The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale. Each fine-scale sequence $y^*_t$ is transformed to $\z$ using Equation (\ref{eqn:z}) and $\Z_{t,t^*}$ is modelled with a CarHMM.
%We assume $\Zone_{t,t^*}$ is Normally distributed with auto-correlation while $\Ztwo_{t,t^*}$ is Gamma distributed without auto-correlation. We also assume that $\Ztwo_{t,t^*}$ and all components of $\Zone_{t,t^*}$ are each independent of one another when conditioned on the subdive state. The emission distribution of each subdive state does not depend upon the dive type. 

@@

Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $Z_{t,t^*}$ when conditioned on $X^*_{t,t^*} = i^*$. 

@@

We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation. Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model. The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. In addition to the CarHHMM-DFT, we consider three similar variations for comparison. As in the full model, all components of $\Z_{t,t^*}$ are assumed to be independent when conditioned on dive type and subdive state for each of the following models:

@@

    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models both $\Z_{t,t^*}$ and as emissions of a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,t^*}$ as emissions of a CarHMM. However, the ``wiggliness"  $\Ztwo_{t,t^*}$ is omitted from this model altogether.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Z_{t,t^*}$ as emissions of a CarHMM. This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 

@@

Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes no auto-correlation within the fine-scale observations, the CarHHMM does not incorporate the ``wiggliness" $\big(\Ztwo_{t,t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.

@@

The three subdive behaviours have distinct average ``wiggliness" values, as the means corresponding to $\Ztwo_{t,t^*}$ are separated by an order of magnitude for each subdive state (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,t^*}$ and the smallest variance corresponding to $\Zone_{t,t^*}$. It also has the highest auto-correlation in $\Zone_{t,t^*}$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean ``wiggliness" ($\Ztwo_{t,t^*}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1. The auto-correlation of acceleration is also slightly lower than in subdive state 1. We therefore hypothesize that subdive state 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $\Ztwo_{t,t^*}$ and variance of $\Zone_{t,t^*}$ in subdive state 3 are both much higher than in the other two states. The auto-correlation of $\Zone_{t,t^*}$ is also much lower. This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).

@@

We use two visual tools to evaluate the CarHHMM-DFT: pseudoresidual plots and empirical histograms. The pseudoresidual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudoresiduals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals mostly support that this model is well-specified. One exception is $\Ztwo_{t,t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:pseudoresids}). This implies that the true distribution of $\Ztwo_{t,t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study. See Sections 1.4 through 1.6 of the supplementary material for pseudoresidual plots for all observations.

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}. Each observed dive duration is weighted by the decoded probability that it corresponds to a particular dive type. This results in two histograms - one corresponding to dive type 1 and another corresponding to dive type 2. Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT. See Sections 1.4 through 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $\z_{t,t^*}$. Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions. In particular, $\Ztwo_{t,t^*}$ is right-skewed and $\Zone_{t,t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

>>>>

We then select a model corresponding to the fine-scale observations of acceleration. Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$. Although $N^*$ is selected heuristically, we test the validity of this model using techniques in Section \ref{subsec:model_validation}.
%
In contrast to the coarse-scale observations, the fine-scale acceleration data exhibit significant sinusoidal behaviour. Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with window size $h=100$ (two seconds) and maximum frequency $\tilde{\omega}=10$ (5 Hz). 
%We select a window size of two seconds since it appears be the smallest window size that captures at least one full period of each ``wiggle" in the acceleration data. 
We then have that $\z_{t,\tilde t^*} = \left\{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\right\}$, where $\zone_{t,\tilde t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window. Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material). Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.

We then select the specific emission distribution of $\Z_{t,\tilde t^*}$. Given the dive types and subdive states, we assume that $\Ztwo_{t,\tilde t^*}$ and all components of $\Zone_{t,\tilde t^*}$ are independent of one another. To reduce model complexity, we also assume that the three fine-scale emission parameters are shared across the two dive types $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$. This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.

All components of $\Zone_{t,\tilde t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) for all $t$ and $\tilde t^*$. Each component is assumed to have its own mean and variance parameters, but all components have the same auto-correlation parameter. 
%Given that $X^*_{t,\tilde t^*} = i^*$, the corresponding unknown parameters are then $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}_{>0}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
Given that $\tilde X^*_{t,\tilde t^*} = i^*$, $\tilde \Ztwo_{t,\tilde t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$. We do not include $\tilde \Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state. 

@@

%The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale. Each fine-scale sequence $y^*_t$ is transformed to $\z$ using Equation (\ref{eqn:z}) and $\Z_{t,\tilde t^*}$ is modelled with a CarHMM.
%We assume $\Zone_{t,\tilde t^*}$ is Normally distributed with auto-correlation while $\Ztwo_{t,\tilde t^*}$ is Gamma distributed without auto-correlation. We also assume that $\Ztwo_{t,\tilde t^*}$ and all components of $\Zone_{t,\tilde t^*}$ are each independent of one another when conditioned on the subdive state. The emission distribution of each subdive state does not depend upon the dive type. 

@@

Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $\Z_{t,\tilde t^*}$ conditioned on $\tilde X^*_{t,\tilde t^*} = i^*$. 

@@

We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation. Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model. The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. In addition to the CarHHMM-DFT, we consider three similar variations for comparison. As in the full model, all components of $\Z_{t,\tilde t^*}$ are assumed to be independent when conditioned on dive type and subdive state for each of the following models:

@@

    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models both $\Z_{t,\tilde t^*}$ and as emissions of a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,\tilde t^*}$ as emissions of a CarHMM. However, the ``wiggliness"  $\Ztwo_{t,\tilde t^*}$ is omitted from this model altogether.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Z_{t,\tilde t^*}$ as emissions of a CarHMM. This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 

@@

Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes no auto-correlation within the fine-scale observations, the CarHHMM does not incorporate the ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.

@@

The three subdive behaviours have distinct average ``wiggliness" values, as the means corresponding to $\Ztwo_{t,\tilde t^*}$ are separated by an order of magnitude for each subdive state (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest variance corresponding to $\Zone_{t,\tilde t^*}$. It also has the highest auto-correlation in $\Zone_{t,\tilde t^*}$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1. The auto-correlation of acceleration is also slightly lower than in subdive state 1. We therefore hypothesize that subdive state 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $\Ztwo_{t,\tilde t^*}$ and variance of $\Zone_{t,\tilde t^*}$ in subdive state 3 are both much higher than in the other two states. The auto-correlation of $\Zone_{t,\tilde t^*}$ is also much lower. This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).

@@

We use two visual tools to evaluate the CarHHMM-DFT: pseudoresidual plots and empirical histograms. The pseudoresidual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudoresiduals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals mostly support that this model is well-specified. One exception is $\Ztwo_{t,\tilde t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:pseudoresids}). This implies that the true distribution of $\Ztwo_{t,\tilde t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study. See Sections 1.4 through 1.6 of the supplementary material for pseudoresidual plots for all observations.

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}. Each observed dive duration is weighted by the decoded probability that it corresponds to a particular dive type. This results in two histograms - one corresponding to dive type 1 and another corresponding to dive type 2. Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT. See Sections 1.4 through 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $\z_{t,\tilde t^*}$. Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions. In particular, $\Ztwo_{t,\tilde t^*}$ is right-skewed and $\Zone_{t,\tilde t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.
